Stream Mining Algorithms for Sensor Data Classification

### 2015-05-27 latex/paper-draft.tex
Modified report to mention streaming datasets.

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Yue Dong}
\IEEEauthorblockA{Department of Mathematics \\ and Statistics\\
University of Ottawa\\
Ottawa, Ontario\\
ydong029@uottawa.ca}
\and
\IEEEauthorblockN{Nathalie Japkowicz}
\IEEEauthorblockA{School of Electrical Engineering\\ and Computer Science\\
	University of Ottawa\\ Ottawa, Ontario\\ nat@site.uottawa.ca \\
\\
}
\and
\IEEEauthorblockN{Vladimir Pestov}
\IEEEauthorblockA{Department of Mathematics \\ and Statistics\\
	University of Ottawa\\
	Ottawa, Ontario\\
	vpest283@uottawa.ca 
}}

@@

SMNN and SENN satisfy the desirable properties of learning systems for efficient data stream mining proposed by G. Hulten and P. Domingos \cite{Pedro}: they require small constant time per data example; they use fixed amount of main memory, irrespective of the total number of examples; they build models using a single scan over the training data; they generate anytime models independent from the order of the examples; and they have the ability to deal with concept drift.

@@

The VFDT overcomes this limitation by building each node of the decision tree based on a small amount of data instances. According to Pedro \cite{VFDT}, the VFDT works as the following: ``given a stream of examples, the first ones will be used to choose the root test; once the root attribute is chosen, the succeeding examples will be passed down to the corresponding leaves and used to choose the appropriate attributes there, and so on recursively." An illustration of how VFDT is built is shown as Figure~\ref{fig:VFDT}.

@@

@@

Another interesting idea we would like to investigate is the use of autoencoders \cite{Hinton1989} for anomaly detection in streaming data. For a lot of streaming data such as surveillance streams, it is very difficult to obtain the instances of the ``anomaly''  because they can be very rare. For example, consider the idea of a smart watch that monitors heart activity in order to detect heart attacks and automatically contact emergency response units. It might be very hard to train a classifier to recognize such event, since we most likely do not have the data concerning the ``anomalous class'' of heart attacks. Moreover, this system needs to have the ability to distinguish non-life critical noise such as the person starting to run.

We feel that traditional supervised learning algorithms for classification will have a hard time to perform well in this scenario. However, an antoencoder can learn from only one class \cite{Japkowicz95}.  An unsupervised learning method would be more suitable for such tasks \cite{Japkowicz01}. Moreover, autoencoders can handle streaming data since they can learn incrementally. This would help deal with concept drift, which is necessary in this task because a person's regular heart activity will change over the years. For example, their resting heart rate will go up or down over the years depending on their health and levels of exercise.

@@

	\bibitem{Pedro}
	Geoff Hulten and Pedro Domingos. Catching up with the data: research issues in mining data streams. \textit{Proc. of Workshop on Research issues in Data Mining and Knowledge Discovery}, 2001.

>>>>

	% author names and affiliations
	% use a multiple column layout for up to three different
	% affiliations
	\author{\IEEEauthorblockN{Yue Dong}
		\IEEEauthorblockA{Department of Mathematics \\ and Statistics\\
			University of Ottawa\\
			Ottawa, Ontario\\
			ydong029@uottawa.ca}
		\and
		\IEEEauthorblockN{Philippe Paradis}
		\IEEEauthorblockA{School of Mathematics \\ and Statistics\\
			Carleton University\\
			Ottawa, Ontario\\
			philippe.paradis@carleton.ca}
		\and
		\IEEEauthorblockN{Nathalie Japkowicz}
		\IEEEauthorblockA{School of Electrical Engineering\\ and Computer Science\\
			University of Ottawa\\ Ottawa, Ontario\\ nat@site.uottawa.ca \\
			\\
		}}

@@

		SMNN and SENN satisfy the desirable properties of learning systems for efficient data stream mining proposed by G. Hulten and P. Domingos \cite{Domingos}: they require small constant time per data example; they use fixed amount of main memory, irrespective of the total number of examples; they build models using a single scan over the training data; they generate anytime models independent from the order of the examples; and they have the ability to deal with concept drift.

@@

		The VFDT overcomes this limitation by building each node of the decision tree based on a small amount of data instances. According to Domingos \cite{VFDT}, the VFDT works as the following: ``given a stream of examples, the first ones will be used to choose the root test; once the root attribute is chosen, the succeeding examples will be passed down to the corresponding leaves and used to choose the appropriate attributes there, and so on recursively." An illustration of how VFDT is built is shown as Figure~\ref{fig:VFDT}.

@@

		\section{Streaming Context}

		NOTE: We might try to merge this with the previous sections so that it looks "seamless", but I decided to write it separately since it is simpler and less confusing.

		NOTE: Need to do a section with and withotu concept drift.

		In the previous section, we constructed the SMNN and the SENN algorithms in the batch learning context. By this, we mean that in this context, the entire dataset is available, it is practical to hold it entirely in memory and the entire dataset is partitioned into mutually exclusive training and testing sets.

		However, SMNN and SENN are stream mining algorithms -- the above was \emph{the first step} in the construction. In this section, we use much larger datasets from real-world sources and show SMNN and SENN work in the stream mining context.

		\subsection{Training and testing sets in the streaming context}

		In the stream mining context, we make the following assumptions:
		\begin{enumerate}
			\item A fraction $\alpha \geq 0$ of the dataset is static and can be pre-loaded, call it $X_\alpha$. This subset can be used for pre-training ()whenever $\alpha > 0$).
			\item Rows that are outside o
			\item A fraction $\beta \geq \alpha$ of the dataset, called $X_\beta$, is labeled ($\beta = 1$ is possible).
			\item The rows in $X_\alpha$ ...
			\item The rows in $X_\beta$ can only be accessed in a single pass (to be more precise, the sliding windows --- which contain a set of rows in $X_\beta$ --- can only be accessed in a single pass)
			\item The memory available is limited such that it is only possible to hold at most a fraction $\gamma \ll 1$ of the dataset in memory (in practice, the value of $\gamma$ depends entirely on
			specifics of the model, including scaling coefficients and constant memory used -- but we ignore such matters for simplicity's sake).
		\end{enumerate}
		Moreover, we assume that the dataset is \emph{ordered}, such that all rows belong to the fraction $\alpha$ described in 1) appear at the beginning of the dataset, before any other row which does not belong to 1). Similarly, the fraction $\beta$ of rows described by 2) appear at the beginning of the dataset.

@@

		%Another interesting idea we would like to investigate is the use of autoencoders \cite{Hinton1989} for anomaly detection in streaming data. For a lot of streaming data such as surveillance streams, it is very difficult to obtain the instances of the ``anomaly''  because they can be very rare. For example, consider the idea of a smart watch that monitors heart activity in order to detect heart attacks and automatically contact emergency response units. It might be very hard to train a classifier to recognize such event, since we most likely do not have the data concerning the ``anomalous class'' of heart attacks. Moreover, this system needs to have the ability to distinguish non-life critical noise such as the person starting to run.

		%We feel that traditional supervised learning algorithms for classification will have a hard time to perform well in this scenario. However, an antoencoder can learn from only one class \cite{Japkowicz95}.  An unsupervised learning method would be more suitable for such tasks \cite{Japkowicz01}. Moreover, autoencoders can handle streaming data since they can learn incrementally. This would help deal with concept drift, which is necessary in this task because a person's regular heart activity will change over the years. For example, their resting heart rate will go up or down over the years depending on their health and levels of exercise.

@@

			\bibitem{Domingos}
			Geoff Hulten and Pedro Domingos. Catching up with the data: research issues in mining data streams. \textit{Proc. of Workshop on Research issues in Data Mining and Knowledge Discovery}, 2001.
